---
title: "Assessing Spatial Heterogeneity in Predicted Whiff Rate Using Geographically Weighted Regression"
author: "Riley Leonard"
date: "April 2023"
output: pdf_document
header-includes:
     - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stargazer)
library(car)
library(GGally)
library(kableExtra)
library(pROC)
library(MASS)
library(glmnet)
library(raster)
library(sp)
library(spData)
library(spgwr)
```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Load Data (Kaggle)

pitches <- read_csv("Statcast_2021.csv")


# Data Wrangling

pitches <- pitches %>%
  dplyr::select(pitch_type, pitch_name, release_speed, release_pos_x, release_pos_z, events, description, 
         zone, stand, p_throws, balls, strikes, bb_type, pfx_x, pfx_z, plate_x, plate_z, outs_when_up, 
         inning, inning_topbot, on_3b, on_2b, on_1b, hc_x, hc_y, vx0, vy0, vz0, ax, ay, az, effective_speed,
         release_spin_rate, release_extension, release_pos_y, estimated_woba_using_speedangle, spin_axis,
         delta_run_exp)

pitches_clean <- pitches %>%
  drop_na(release_speed, release_pos_x, release_pos_z, pfx_x, pfx_z, 
          plate_x, plate_z, vx0, vy0, vz0, ax, ay, az, effective_speed,
          release_spin_rate, release_extension, release_pos_y, spin_axis)


# Filter Swings

swings <- pitches_clean %>%
  filter(description %in% c("hit_into_play", "swinging_strike_blocked", "foul", "foul_tip","swinging_strike"))


# Filter by Handedness

swingsRHB <- swings %>%
  filter(stand == "R")

swingsLHB <- swings %>%
  filter(stand == "L")


# Filter by Pitch Type

swingsFastRHB<- swingsRHB %>%
  filter(pitch_type %in% c("FF", "SI"))

swingsOffRHB <- anti_join(swingsRHB, swingsFastRHB)

swingsFastLHB<- swingsLHB %>%
  filter(pitch_type %in% c("FF", "SI"))

swingsOffLHB <- anti_join(swingsLHB, swingsFastLHB)


# Create Samples

set.seed(0)

sampleFastRHB <- sample_n(swingsFastRHB, size = 15000)

sampleOffRHB <- sample_n(swingsOffRHB, size = 15000)

sampleFastLHB <- sample_n(swingsFastLHB, size = 15000)

sampleOffLHB <- sample_n(swingsOffLHB, size = 15000)


# Write CSVs

write_csv(sampleFastRHB, "swingsFastRHB.csv")

write_csv(sampleOffRHB, "swingsOffRHB.csv")

write_csv(sampleFastLHB, "swingsFastLHB.csv")

write_csv(sampleOffLHB, "swingsOffLHB.csv")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# Load Data

fastRHB <- read_csv("swingsFastRHB.csv")
offRHB <- read_csv("swingsOffRHB.csv")
fastLHB <- read_csv("swingsFastLHB.csv")
offLHB <- read_csv("swingsOffLHB.csv")

pitches <- read_csv("swingData.csv")
```


```{r include=FALSE}
# Wrangling

pitches <- pitches %>%
  mutate(whiff = ifelse(description %in% c("swinging_strike", "swinging_strike_blocked"), 1, 0),
         no_balls = ifelse(balls == 0, 1, 0),
         one_ball = ifelse(balls == 1, 1, 0),
         two_balls = ifelse(balls == 2, 1, 0),
         three_balls = ifelse(balls == 3, 1, 0),
         no_strikes = ifelse(strikes == 0, 1, 0),
         one_strike = ifelse(strikes == 1, 1, 0),
         two_strikes = ifelse(strikes == 2, 1, 0)) %>%
  rename(horz_movement = pfx_x,
         vert_movement = pfx_z) %>%
  mutate(horz_movement = abs(horz_movement))
  

fastRHB <- fastRHB %>%
  mutate(whiff = ifelse(description %in% c("swinging_strike", "swinging_strike_blocked"), 1, 0),
         no_balls = ifelse(balls == 0, 1, 0),
         one_ball = ifelse(balls == 1, 1, 0),
         two_balls = ifelse(balls == 2, 1, 0),
         three_balls = ifelse(balls == 3, 1, 0),
         no_strikes = ifelse(strikes == 0, 1, 0),
         one_strike = ifelse(strikes == 1, 1, 0),
         two_strikes = ifelse(strikes == 2, 1, 0)) %>%
  rename(horz_movement = pfx_x,
         vert_movement = pfx_z) %>%
  mutate(horz_movement = abs(horz_movement))


offRHB <- offRHB %>%
  mutate(whiff = ifelse(description %in% c("swinging_strike", "swinging_strike_blocked"), 1, 0),
         no_balls = ifelse(balls == 0, 1, 0),
         one_ball = ifelse(balls == 1, 1, 0),
         two_balls = ifelse(balls == 2, 1, 0),
         three_balls = ifelse(balls == 3, 1, 0),
         no_strikes = ifelse(strikes == 0, 1, 0),
         one_strike = ifelse(strikes == 1, 1, 0),
         two_strikes = ifelse(strikes == 2, 1, 0)) %>%
  rename(horz_movement = pfx_x,
         vert_movement = pfx_z) %>%
  mutate(horz_movement = abs(horz_movement))


fastLHB <- fastLHB %>%
  mutate(whiff = ifelse(description %in% c("swinging_strike", "swinging_strike_blocked"), 1, 0),
         no_balls = ifelse(balls == 0, 1, 0),
         one_ball = ifelse(balls == 1, 1, 0),
         two_balls = ifelse(balls == 2, 1, 0),
         three_balls = ifelse(balls == 3, 1, 0),
         no_strikes = ifelse(strikes == 0, 1, 0),
         one_strike = ifelse(strikes == 1, 1, 0),
         two_strikes = ifelse(strikes == 2, 1, 0)) %>%
  rename(horz_movement = pfx_x,
         vert_movement = pfx_z) %>%
  mutate(horz_movement = abs(horz_movement))



offLHB <- offLHB %>%
  mutate(whiff = ifelse(description %in% c("swinging_strike", "swinging_strike_blocked"), 1, 0),
         no_balls = ifelse(balls == 0, 1, 0),
         one_ball = ifelse(balls == 1, 1, 0),
         two_balls = ifelse(balls == 2, 1, 0),
         three_balls = ifelse(balls == 3, 1, 0),
         no_strikes = ifelse(strikes == 0, 1, 0),
         one_strike = ifelse(strikes == 1, 1, 0),
         two_strikes = ifelse(strikes == 2, 1, 0)) %>%
  rename(horz_movement = pfx_x,
         vert_movement = pfx_z) %>%
  mutate(horz_movement = abs(horz_movement))
```



## Non-uniform Distribution of Response Variable in Space

```{r fig.height=5, fig.width=4, message=FALSE, warning=FALSE}
# Accuracy Plot

topKzone = 3.5
botKzone = 1.6
inKzone = -.95
outKzone = 0.95
kZone = data.frame(
  x = c(inKzone, inKzone, outKzone, outKzone, inKzone)
  , y = c(botKzone, topKzone, topKzone, botKzone, botKzone)
)


pitches %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location") +
  geom_path(aes(x, y), data = kZone) +
  theme_classic()

pitches %>%
  filter(whiff == 0) %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location") +
  geom_path(aes(x, y), data = kZone) +
  theme_classic()

pitches %>%
  filter(whiff == 1) %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location") +
  geom_path(aes(x, y), data = kZone) +
  theme_classic()

pitches %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
    stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
    scale_fill_distiller(palette= "Spectral", direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location") +
  geom_path(aes(x, y), data = kZone) +
  theme_classic()

pitches %>%
  filter(whiff == 1) %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
    stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
    scale_fill_distiller(palette= "Spectral", direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location") +
  geom_path(aes(x, y), data = kZone) +
  theme_classic()
```

```{r fig.height=10, fig.width=8, message=FALSE, warning=FALSE}
# Density Plots

offRHB %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Density") +
  ggtitle("Pitches (RHB)") +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))

offRHB %>%
  filter(whiff == 0) %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Density") +
  ggtitle("Non-whiffs (RHB)") +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))

offRHB %>%
  filter(whiff == 1) %>%
  ggplot(aes(x = plate_x, y = plate_z)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Density") +
  ggtitle("Whiffs (RHB)") +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))
```

*Conclusion: fitted coefficient values of a global parametric model may fail to capture local variations in the data.*

## Analyzing Spatial Autocorrelation

```{r}
# Spatial Autocorrelation

# Create Polygons

# Define Adjacent Polygons
```



## Parametric Model for Whiff Rate

```{r message=FALSE, warning=FALSE}
## Create Parametric Model for Whiff Rate

set.seed(0)


# Create Training and Test Data

Train <- pitches %>% sample_frac(2/3)
Test <- anti_join(pitches, Train)


# Logistic Regression

logistic_model <- glm(whiff ~ release_speed + release_spin_rate + horz_movement + vert_movement + ay +
                      one_ball + two_balls + three_balls + one_strike + two_strikes,
                      data = Train, family = "binomial")


# Logistic Regression Fastballs RHB

TrainFastRHB <- fastRHB %>% sample_frac(2/3)
TestFastRHB<- anti_join(pitches, TrainFastRHB)

log_fast_rhb <- glm(whiff ~ release_speed + release_spin_rate + horz_movement + vert_movement + ay,
                    data = TrainFastRHB, family = "binomial")


# Logistic Regression Off-Speeds RHB

TrainOffRHB <- offRHB %>% sample_frac(2/3)
TestOffRHB<- anti_join(pitches, TrainOffRHB)


log_off_rhb <- glm(whiff ~ release_speed + release_spin_rate + horz_movement + vert_movement + ay,
                   data = TrainOffRHB, family = "binomial")


```


Important differences between models:

* In general model, the coefficient estimate for release speed is negative, since it is confounded by pitch type
* In the pitch-specific models, the coefficient estimate for release speed is positive but non-significant
* The estimated coefficient for the vertical movement parameter is significant and negative for off-speed pitches (more downward movement is associated with a greater whiff probability)
* The estimated coefficient for the vertical movement parameter is significant and positive for fastballs (less downward movement is associated with a greater whiff probability)
* The estimated coefficient for the horizontal movement parameter is significant and negative for fastballs
* Acceleration is associated with a significant positive increase in whiff probability for both pitch families
* The count indicators are mostly significant and exist to control for contextual confounds and provide model stabilizing effects. The count indicators improve predictive performance, but do not meaningfully effect the direction and magnitude of the coefficient estimates of interest (they will thus be omitted from the GWR).



### GWR (Off-speed RHB)

```{r}
# Model Predictions

pred <- predict(log_off_rhb, offRHB, type = "response")
data <- cbind(offRHB, pred)
```


## Covriate Raster (predictions)

```{r}
# Covariate Raster

data_short <- offRHB %>%
  dplyr::select(release_speed, release_spin_rate, horz_movement, vert_movement, ay)

pts <- SpatialPoints(c(offRHB[ ,16], offRHB[ ,17]))

spData <- SpatialPointsDataFrame(pts, data = data_short)

r <- raster(spData)
res(r) <- 0.1
f <- rasterize(spData, r)
plot(f)
```


## Covariate Raster (new)

```{r}
# Covariate Raster

data_short2 <- data %>%
  dplyr::select(pred)

spData2 <- SpatialPointsDataFrame(pts, data = data_short2)

r2 <- raster(spData2)
res(r2) <- 0.1
f2 <- rasterize(spData2, r2)
plot(f2) +
  geom_path(aes(x, y), data = kZone)
```


## GWR

```{r}
set.seed(0)

pred <- predict(log_off_rhb, TrainOffRHB, type = "response")

data_sample_short <- TrainOffRHB %>%
  dplyr::select(release_speed, release_spin_rate, horz_movement, vert_movement, ay)

data_sample_short <- cbind(data_sample_short, pred)

coords <- data.frame(x = TrainOffRHB[ ,16], y = TrainOffRHB[ ,17])

sample_pts <- SpatialPoints(coords)

spSample <- SpatialPointsDataFrame(sample_pts, data = data_sample_short)

r2 <- raster(spSample)

res(r2) <- 0.1

fitpoints <- rasterToPoints(r2)

#GWRbandwidth <- gwr.sel(pred ~ release_speed + release_spin_rate + horz_movement + vert_movement + ay, 
#                        data = spSample, coords = coords, adapt = T) 

gwr.model <- gwr(pred ~ release_speed + release_spin_rate + horz_movement + vert_movement + ay,
                 data = spSample,
                 longlat = FALSE,
                 adapt = 0.01121417,
                 fit.points = fitpoints)
```

```{r}
sp <- gwr.model$SDF
spplot(sp, 'sum.w')
```

```{r fig.height=10, fig.width=8, message=FALSE, warning=FALSE}
cells <- cellFromXY(r2, fitpoints)
dd <- as.matrix(data.frame(sp))
b <- brick(r2, values = FALSE, nl = ncol(dd))
b[cells] <- dd
names(b) <- colnames(dd)
plot(b)

results <- as.data.frame(gwr.model$SDF)

results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = vert_movement)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Vertical Movement") +
  scale_fill_distiller(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))

results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = release_speed)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Release Speed") +
  scale_fill_distiller(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))

results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = release_spin_rate)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Spin Rate") +
  scale_fill_distiller(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))


results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = horz_movement)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Horizontal Movement") +
  scale_fill_distiller(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))


results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = vert_movement)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Vertical Movement") +
  scale_fill_fermenter(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))


results %>%
  ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = ay)) +
    labs(y = "Vertical Location",
         x = "Horizontal Location",
         fill = "Parameter Estimate") +
  ggtitle("Acceleration") +
  scale_fill_distiller(direction = 1) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 5)) +
  geom_path(aes(x, y), data = kZone, linewidth = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 24),
        axis.title.y = element_text(size = 28, vjust = 4),
        axis.title.x = element_text(size = 28, vjust = -1),
        legend.text = element_text(size = 24),
        legend.title = element_text(size = 28, vjust = 3),
        legend.key.size = unit(1.5, 'cm'),
        plot.margin = ggplot2::margin(18, 18, 18, 18, 'points'),
        plot.title = element_text(size = 32, face = "bold.italic", hjust = 0.5))

```

*Isotropic spatial weights kernel bandwidth of 0.01121417 determined by leave-one-out cross-validation.*


## New Scatter Plot with Color by Predicted Whiff Rate


## Extensions

Model for continuous response (xWOBA)




